---
title: "BMD Parameterisation in BMABMDR: Design Choices, Consequences, and the Prior on BMD"
author: "Zhenglei Gao <zhenglei.gao@bayer.com>"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(BMABMDR)
library(dplyr)
library(ggplot2)
library(drc)
library(bmd)

# Helper: format tibble as kable for Word-friendly output
ktbl <- function(x, ...) knitr::kable(x, format = "pipe", ...)
```

```{r}
packageVersion("BMABMDR")
```

# Overview

This document examines a structural design choice in BMABMDR: the BMR value `q`
is fixed data baked into each Stan model at the data-preparation stage, making
the **BMD the primary sampled parameter** (`par2`) rather than a derived
quantity. We document the consequences of this choice, characterise the prior
placed on BMD, evaluate whether that prior is truly non-informative, and
demonstrate the practical impact by comparing BMD10/BMD20/BMD50 estimates
obtained via three separate runs versus estimates derived post-hoc from a single
BMD10 posterior.

---

# The parameterisation

## The Exponential (E4) dose-response model

The dose-response mean in the E4 Normal model is:

$$\mu(x) = a \left[1 + (c-1)\left(1 - e^{-b\, x^d}\right)\right]$$

with parameters:

| Symbol | Stan name | Interpretation |
|--------|-----------|----------------|
| $a$    | `par1`    | background response |
| $\text{BMD}_q$ | `par2` | dose at which the relative change equals $q$ |
| $c$    | `par3`    | fold change (asymptote / background), so $\mu_\infty = c \cdot a$ |
| $d$    | `exp(par4)` | Hill slope / power |
| $\sigma^2$ | `exp(-par5)` | residual variance |

The slope parameter $b$ is **not sampled**. It is computed in Stan's
`transformed parameters` block by solving $\mu(\text{BMD}_q) - \mu(0) =
q \cdot \mu(0)$ for $b$:

$$b = e^{-\log(\text{BMD}_q)\cdot d}\cdot\left[-\log\!\left(1 - \frac{q}{c-1}\right)\right]$$

This reparameterisation is exact — $b$ is a deterministic function of
`(par2, par3, par4, q)`. It makes `par2` directly interpretable as BMD, but
it **encodes `q` as a compile-time constant in the data block**, not as a
parameter.

## Consequence: one posterior per BMR level

Because `q` is fixed data, a separate call to `PREP_DATA_N()` and
`full.laplace_MA()` / `sampling_MA()` is needed for each desired BMR level.
The 16 models are refitted from scratch each time.

More importantly, the **posterior model weights differ across runs** because:

1. The PERT prior on `par2` has data-adaptive hyperparameters estimated via
   GAMLSS for each specific `q` (see Section 3).
2. Marginal likelihoods are computed under that `q`-specific prior, so model
   weights can change when `q` changes.

The BMA-BMD10, BMA-BMD20, and BMA-BMD50 therefore come from **three different
model-averaged posteriors**, not one coherent joint posterior.

---

# The prior on BMD (`par2`)

## Prior family

The prior on `par2` is a **PERT (Power PERT) distribution** — a rescaled Beta
distribution — supported on $[\text{BMD}_{\min},\, \text{BMD}_{\max}]$ with
mode $\text{BMD}_{\text{mode}}$ and shape parameter $g$ (`shape.BMD`):

$$\alpha = 1 + g \cdot \frac{\text{BMD}_{\text{mode}} - \text{BMD}_{\min}}
                             {\text{BMD}_{\max}  - \text{BMD}_{\min}}, \qquad
  \beta  = 1 + g \cdot \frac{\text{BMD}_{\max}  - \text{BMD}_{\text{mode}}}
                             {\text{BMD}_{\max}  - \text{BMD}_{\min}}$$

When $g \to 0$ the PERT approaches $\text{Beta}(1, 1)$, i.e. a uniform on the
support — the "non-informative" limit.

## Default hyperparameter values

From `PREP_DATA_N()` (on the rescaled dose scale $x \in [0, 1]$):

| Hyperparameter | Default value | Source |
|---|---|---|
| `BMD.min` | `.Machine$double.xmin` $\approx 0$ | hard-coded |
| `BMD.max` | `3` (if `extended = TRUE`) | hard-coded |
| `BMD.mode` | `0.5` | hard-coded |
| `shape.BMD` | `0.0001` | hard-coded default argument |

With `shape.BMD = 0.0001` the PERT is numerically indistinguishable from
$\text{Uniform}(0, 3)$ on the rescaled dose scale. The package documentation
describes this as "non-informative".

## Is it truly non-informative?

### The uniform is not non-informative under transformation

The prior is uniform on the **rescaled** BMD, i.e. $\text{BMD} / \max(\text{dose})$.
On the **original dose scale** the prior is:

$$\text{par2} \sim \text{Uniform}(0,\; 3 \cdot \max(\text{dose}))$$

This is data-adaptive: the upper bound depends on `max(dose)`. Two analysts
using datasets with different maximum doses get different priors on the same
physical BMD. This violates the spirit of a prior specified independently of
the data.

### The GAMLSS start value is data-driven — but distinct from the prior

It is important to distinguish two quantities that are both derived from the
GAMLSS fractional-polynomial fit:

- **`bmd.sv`** — the HMC *initial value* for `par2`. Stan starts the Markov
  chain here. It does **not** appear in the prior or likelihood; it only
  affects how quickly the chain reaches the high-density region. A well-mixed
  chain converges to the same posterior distribution regardless of the start
  value. `bmd.sv` is always data-driven (GAMLSS estimate of the dose
  achieving BMR `q`), but this does not make the prior informative.

- **`BMD.mode`** — the *mode* of the PERT prior, passed to Stan as
  `priormu[2]`. In the default case this is hard-coded to `0.5` on the
  rescaled dose scale, regardless of the data. The PERT shape `0.0001` then
  makes the prior nearly flat across the whole support, so the mode is
  practically irrelevant.

The flat prior is therefore **not compromised by the data-driven start
value**. The two only coincide when the user explicitly supplies `prior.BMD`:
in that case `BMD.mode` is set to the user-specified mode (on the original
dose scale, then rescaled), making the prior genuinely informative. The
GAMLSS estimate also sets the initial value to the prior mode in this case,
but that is incidental — it is `BMD.mode` entering Stan's `model` block that
creates the informative prior, not `bmd.sv`.

### The effective prior range depends on `extended`

The upper bound `BMD.max = 3` (when `extended = TRUE`) means the prior allows
the BMD to be up to three times the maximum observed dose. This is a
substantive assumption: it implies the BMD is unlikely to lie above
$3 \cdot \max(\text{dose})$. However, in practice, we also do not want to estimate a BMD beyond the tested dose range. 

### Conclusion on non-informativeness

The prior on `par2` is:

- **Approximately uniform** on $[0, 3 \cdot \max(\text{dose})]$ by default
  (shape parameter 0.0001)
- **Not invariant to reparameterisation** — a uniform on BMD is not uniform
  on $\log(\text{BMD})$ or on $b$
- **Data-adaptive in its support** — the upper bound scales with `max(dose)`
- **`q`-specific** — the GAMLSS start value and any user-supplied `prior.BMD`
  are computed for a specific BMR level

The claim "non-informative prior on BMD10 / BMD20" is therefore only
approximately true in a narrow operational sense: the PERT shape is nearly
flat. The broader sense — prior independence from data and BMR choice — does
not hold.

---

# Demonstration: BMD10, BMD20, BMD50 from separate runs

We use `immunotoxicityData` (continuous, Normal, 5 dose groups) throughout.




```{r load-data}
data("immunotoxicityData")
# PREP_DATA_N with sumstats = TRUE expects exactly one row per dose group.
# The package examples use rows 1:5 (one female experiment, 5 dose levels:
# 0, 0, 0.475, 4.75, 47.5 — note two dose-0 controls).
dat <- as.data.frame(immunotoxicityData[1:5, c("Dose", "Mean", "SD", "n")])
ktbl(dat, caption = "Immunotoxicity data (rows 1–5)")
```

## Prepare data for each BMR level

```{r prep-data}
data_N10 <- PREP_DATA_N(data = dat, sumstats = TRUE, sd = TRUE, q = 0.10)
data_N20 <- PREP_DATA_N(data = dat, sumstats = TRUE, sd = TRUE, q = 0.20)
data_N50 <- PREP_DATA_N(data = dat, sumstats = TRUE, sd = TRUE, q = 0.50)

data_LN10 <- PREP_DATA_LN(data = dat, sumstats = TRUE, sd = TRUE, q = 0.10)
data_LN20 <- PREP_DATA_LN(data = dat, sumstats = TRUE, sd = TRUE, q = 0.20)
data_LN50 <- PREP_DATA_LN(data = dat, sumstats = TRUE, sd = TRUE, q = 0.50)
```

## Fit all three BMR levels with HMC (bridge sampling)

We use `sampling_MA()` so that full MCMC draws are available for the
post-hoc inversion in Section 5.

```{r fit-sampling, cache=TRUE}
fit10 <- sampling_MA(data_N10, data_LN10, prior.weights = rep(1, 16))
fit20 <- sampling_MA(data_N20, data_LN20, prior.weights = rep(1, 16))
fit50 <- sampling_MA(data_N50, data_LN50, prior.weights = rep(1, 16))
```

## Fit all three BMR levels with Laplace approximation

`full.laplace_MA()` uses MAP + Hessian approximation (no HMC draws). Weights
are based on the Laplace-approximated marginal likelihood.

```{r fit-laplace, cache=TRUE}
lp10 <- full.laplace_MA(data_N10, data_LN10, prior.weights = rep(1, 16))
lp20 <- full.laplace_MA(data_N20, data_LN20, prior.weights = rep(1, 16))
lp50 <- full.laplace_MA(data_N50, data_LN50, prior.weights = rep(1, 16))
```

## Extract BMA results

### Bridge-sampling (HMC) estimates

The `BMDMixtureBS` component contains draws from the model-averaged posterior
weighted by bridge-sampling posterior model probabilities.

```{r extract-direct}
bmd_bs <- bind_rows(
  tibble(BMR = "10%", Method = "Bridge Sampling",
         BMDL = quantile(fit10$BMDMixtureBS, 0.05),
         BMD  = quantile(fit10$BMDMixtureBS, 0.50),
         BMDU = quantile(fit10$BMDMixtureBS, 0.95)),
  tibble(BMR = "20%", Method = "Bridge Sampling",
         BMDL = quantile(fit20$BMDMixtureBS, 0.05),
         BMD  = quantile(fit20$BMDMixtureBS, 0.50),
         BMDU = quantile(fit20$BMDMixtureBS, 0.95)),
  tibble(BMR = "50%", Method = "Bridge Sampling",
         BMDL = quantile(fit50$BMDMixtureBS, 0.05),
         BMD  = quantile(fit50$BMDMixtureBS, 0.50),
         BMDU = quantile(fit50$BMDMixtureBS, 0.95))
)
```

### Laplace approximation estimates

```{r extract-laplace}
bmd_lp <- bind_rows(
  tibble(BMR = "10%", Method = "Laplace",
         BMDL = lp10$MA[1], BMD = lp10$MA[2], BMDU = lp10$MA[3]),
  tibble(BMR = "20%", Method = "Laplace",
         BMDL = lp20$MA[1], BMD = lp20$MA[2], BMDU = lp20$MA[3]),
  tibble(BMR = "50%", Method = "Laplace",
         BMDL = lp50$MA[1], BMD = lp50$MA[2], BMDU = lp50$MA[3])
)
```

### Laplace vs Bridge Sampling: BMD comparison

```{r bmd-lp-vs-bs}
bmd_all <- bind_rows(bmd_bs, bmd_lp) |>
  mutate(across(c(BMDL, BMD, BMDU), ~ round(.x, 3)))

bmd_all |>
  tidyr::pivot_wider(
    names_from  = Method,
    values_from = c(BMDL, BMD, BMDU),
    names_glue  = "{Method}_{.value}"
  ) |>
  dplyr::select(BMR,
         `Laplace_BMDL`, `Bridge Sampling_BMDL`,
         `Laplace_BMD`,  `Bridge Sampling_BMD`,
         `Laplace_BMDU`, `Bridge Sampling_BMDU`)
```

## Compare posterior model weights across BMR levels

```{r compare-weights}
# weights_bridge_sampling is a named numeric vector of length 16
wts <- bind_rows(
  tibble(Model = names(fit10$weights_bridge_sampling),
         weight = unname(fit10$weights_bridge_sampling),
         BMR = "10%"),
  tibble(Model = names(fit20$weights_bridge_sampling),
         weight = unname(fit20$weights_bridge_sampling),
         BMR = "20%"),
  tibble(Model = names(fit50$weights_bridge_sampling),
         weight = unname(fit50$weights_bridge_sampling),
         BMR = "50%")
)

wts |>
  tidyr::pivot_wider(names_from = BMR, values_from = weight) |>
  arrange(desc(`10%`))
```

Differences in this table confirm that the three BMA results are not based on
the same model-averaged posterior.

---

# Post-hoc inversion: deriving multiple BMDs from a single HMC run

## Approach

From a `sampling_MA()` run fitted at `q = 0.10`, each MCMC draw contains
`(par1, par2, par3, par4, par5)` for each model. Because `b` is a
deterministic function of these parameters and the original `q = 0.10`, we can
solve the dose-response equation for any other BMR level $q'$ analytically.

For the Exponential (E4, Normal) model the BMD at BMR $q'$ is:

$$\text{BMD}_{q'} = \exp\!\left(\frac{1}{d}\log\!\left(\frac{-\log(1 - q'/(c-1))}{b}\right)\right)$$

where $b$ is recovered from the stored parameters and the fitted `q = 0.10`.

**Note:** This inversion must be implemented separately for each of the 16
model shapes — each has a different algebraic form. For the Laplace
approximation (`full.laplace_MA`) only the MAP estimate and Hessian are stored,
not full posterior draws, so this post-hoc inversion is not directly feasible
without additional sampling.

## Inversion helper

```{r inversion-helpers}
# fit$parsN is a list of 8 data frames (one per Normal model shape, in order:
# E4, IE4, H4, LN4, G4, QE4, P4, L4). Each has columns a, b, c, d, BMD, ...
# b is stored directly from Stan's transformed parameters block.
# Inverting the E4 dose-response for a new BMR q_new (increasing Normal):
#   BMD(q') = exp( log( -log(1 - q'/(c-1)) / b ) / d )

bmd_from_draws <- function(draws_df, q_new, maxD) {
  b   <- draws_df$b
  c_r <- draws_df$c
  d   <- draws_df$d
  log_arg <- -log(1 - q_new / (c_r - 1)) / b
  valid   <- is.finite(log_arg) & log_arg > 0
  bmd_new <- rep(NA_real_, nrow(draws_df))
  bmd_new[valid] <- exp(log(log_arg[valid]) / d[valid]) * maxD
  bmd_new
}
```

## Single-model comparison: E4_N at BMD10, BMD20, BMD50

To make a fair comparison we use the **E4_N model only** across all three
dedicated runs (BMD10, BMD20, BMD50), and compare those direct per-run
estimates to the post-hoc inversion of the BMD10 posterior into BMD20 and
BMD50. All estimates are from the same model, so they are directly comparable.

```{r single-model-comparison}
maxD <- data_N10$data$maxD

# Direct per-run E4_N estimates: BMD column is on rescaled scale in parsN[[1]]
direct_E4N <- bind_rows(
  tibble(BMR = "10%", source = "direct",
         BMDL = quantile(fit10$parsN[[1]]$BMD * maxD, 0.05, na.rm = TRUE),
         BMD  = quantile(fit10$parsN[[1]]$BMD * maxD, 0.50, na.rm = TRUE),
         BMDU = quantile(fit10$parsN[[1]]$BMD * maxD, 0.95, na.rm = TRUE)),
  tibble(BMR = "20%", source = "direct",
         BMDL = quantile(fit20$parsN[[1]]$BMD * data_N20$data$maxD, 0.05, na.rm = TRUE),
         BMD  = quantile(fit20$parsN[[1]]$BMD * data_N20$data$maxD, 0.50, na.rm = TRUE),
         BMDU = quantile(fit20$parsN[[1]]$BMD * data_N20$data$maxD, 0.95, na.rm = TRUE)),
  tibble(BMR = "50%", source = "direct",
         BMDL = quantile(fit50$parsN[[1]]$BMD * data_N50$data$maxD, 0.05, na.rm = TRUE),
         BMD  = quantile(fit50$parsN[[1]]$BMD * data_N50$data$maxD, 0.50, na.rm = TRUE),
         BMDU = quantile(fit50$parsN[[1]]$BMD * data_N50$data$maxD, 0.95, na.rm = TRUE))
)

# Post-hoc inversion from the BMD10 E4_N posterior
bmd20_posthoc <- bmd_from_draws(fit10$parsN[[1]], q_new = 0.20, maxD = maxD)
bmd50_posthoc <- bmd_from_draws(fit10$parsN[[1]], q_new = 0.50, maxD = maxD)

posthoc_E4N <- bind_rows(
  tibble(BMR = "20%", source = "post-hoc from BMD10",
         BMDL = quantile(bmd20_posthoc, 0.05, na.rm = TRUE),
         BMD  = quantile(bmd20_posthoc, 0.50, na.rm = TRUE),
         BMDU = quantile(bmd20_posthoc, 0.95, na.rm = TRUE)),
  tibble(BMR = "50%", source = "post-hoc from BMD10",
         BMDL = quantile(bmd50_posthoc, 0.05, na.rm = TRUE),
         BMD  = quantile(bmd50_posthoc, 0.50, na.rm = TRUE),
         BMDU = quantile(bmd50_posthoc, 0.95, na.rm = TRUE))
)

bind_rows(direct_E4N, posthoc_E4N) |>
  arrange(BMR, source) |>
  ktbl(caption = "E4_N model: direct vs post-hoc BMD estimates", digits = 3)
```

## Visualisation

```{r plot-comparison, fig.width=8, fig.height=5}
bind_rows(direct_E4N, posthoc_E4N) |>
  mutate(BMR = factor(BMR, levels = c("10%", "20%", "50%"))) |>
  ggplot(aes(x = source, y = BMD, ymin = BMDL, ymax = BMDU,
             colour = source)) +
  geom_pointrange(size = 0.8, position = position_dodge(0.3)) +
  facet_wrap(~BMR, scales = "free_y") +
  labs(
    title    = "E4_N model: direct run vs post-hoc inversion from BMD10",
    subtitle = "Direct: dedicated sampling_MA() run per BMR; Post-hoc: b-inversion of BMD10 draws",
    x = NULL, y = "BMD (original dose scale)",
    colour = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom", axis.text.x = element_blank())
```

## Posterior weights across BMR levels

```{r weights-table}
tibble(
  Model = names(fit10$weights_bridge_sampling),
  `BMD10` = round(fit10$weights_bridge_sampling, 4),
  `BMD20` = round(fit20$weights_bridge_sampling, 4),
  `BMD50` = round(fit50$weights_bridge_sampling, 4)
) |>
  arrange(desc(BMD10)) |>
  ktbl(caption = "Bridge Sampling posterior model weights across BMR levels")
```

## Posterior parameter estimates for E4_N

Median and 95% credible interval for each parameter, across the three runs.
Because `b` is derived from `(BMD, c, d, q)`, its value changes across runs
even when the underlying dose-response curve is the same — reflecting the
different BMR definitions.

```{r param-table}
summarise_pars <- function(pars_df, bmr_label, maxD) {
  fmt <- function(x, digits = 3) round(x, digits)
  ci  <- function(x, digits = 3)
    paste0("[", fmt(quantile(x, 0.025), digits), ", ",
                fmt(quantile(x, 0.975), digits), "]")

  tibble(
    BMR       = bmr_label,
    Parameter = c("a (med)", "a (95% CI)",
                  "b (med)", "b (95% CI)",
                  "c (med)", "d (med)",
                  "BMD (med)", "BMD (95% CI)"),
    Value = c(
      as.character(fmt(median(pars_df$a))),         ci(pars_df$a),
      as.character(fmt(median(pars_df$b), 4)),      ci(pars_df$b, 4),
      as.character(fmt(median(pars_df$c))),
      as.character(fmt(median(pars_df$d))),
      as.character(fmt(median(pars_df$BMD * maxD))), ci(pars_df$BMD * maxD)
    )
  )
}

bind_rows(
  summarise_pars(fit10$parsN[[1]], "10%", data_N10$data$maxD),
  summarise_pars(fit20$parsN[[1]], "20%", data_N20$data$maxD),
  summarise_pars(fit50$parsN[[1]], "50%", data_N50$data$maxD)
) |>
  tidyr::pivot_wider(names_from = BMR, values_from = Value) |>
  ktbl(caption = "E4_N posterior parameter estimates across BMR levels")
```

## Do individual model estimates change across parameterisations?

The model-averaged BMD differs across BMR parameterisations partly because
the **posterior model weights** shift. But a second, more fundamental source
of variation is that the **individual model parameter estimates themselves
change**. This is expected: although the data and likelihood are identical in
each run, the Stan model samples `par2 = BMD_q` directly — a different
physical quantity for each `q` — and derives `b` from `(par2, par3, par4, q)`.
The prior on `par2` is always PERT(0, 3) on the rescaled scale, but it
implies a **different prior on the underlying curve shape** for each `q`.
This propagates through the joint posterior, shifting the marginal
distributions of all parameters, not just BMD.

The table below compares the posterior medians of the "invariant" curve-shape
parameters (`a`, `c`, `d`) and the derived slope `b` across all eight Normal
models and the three BMR parameterisations. If the parameterisation were
irrelevant, columns BMD10/BMD20/BMD50 would be identical for each parameter.

```{r param-invariance, fig.width=9, fig.height=6}
model_names_N <- c("E4_N", "IE4_N", "H4_N", "LN4_N",
                   "G4_N", "QE4_N", "P4_N", "L4_N")

# Extract median of a, b, c, d for each model × BMR
extract_medians <- function(fit_obj, bmr_label) {
  purrr::map2_dfr(fit_obj$parsN, model_names_N, function(df, nm) {
    tibble(
      Model = nm,
      BMR   = bmr_label,
      a_med = median(df$a, na.rm = TRUE),
      b_med = median(df$b, na.rm = TRUE),
      c_med = median(df$c, na.rm = TRUE),
      d_med = median(df$d, na.rm = TRUE)
    )
  })
}

par_comparison <- bind_rows(
  extract_medians(fit10, "BMD10"),
  extract_medians(fit20, "BMD20"),
  extract_medians(fit50, "BMD50")
)

# Wide table: one row per model, columns for each param × BMR
par_wide <- par_comparison |>
  tidyr::pivot_wider(
    names_from  = BMR,
    values_from = c(a_med, b_med, c_med, d_med),
    names_glue  = "{BMR}_{.value}"
  ) |>
  dplyr::select(
    Model,
    BMD10_a = BMD10_a_med, BMD20_a = BMD20_a_med, BMD50_a = BMD50_a_med,
    BMD10_c = BMD10_c_med, BMD20_c = BMD20_c_med, BMD50_c = BMD50_c_med,
    BMD10_d = BMD10_d_med, BMD20_d = BMD20_d_med, BMD50_d = BMD50_d_med,
    BMD10_b = BMD10_b_med, BMD20_b = BMD20_b_med, BMD50_b = BMD50_b_med
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

ktbl(par_wide,
     caption = "Posterior median of curve parameters across BMR parameterisations (8 Normal models)")
```

The derived slope `b` is expected to differ (it absorbs the BMR definition).
But `a` (background), `c` (fold change), and `d` (Hill power) should in
principle be invariant properties of the dose-response curve. Any differences
in these columns are caused by the `q`-specific prior acting through `b` and
propagating into the joint posterior via correlations between parameters.

To quantify the magnitude of these shifts, we compute the range of each
parameter's posterior median across BMR levels, expressed as a percentage of
the BMD10 value:

```{r param-shift-magnitude}
shift_summary <- par_comparison |>
  tidyr::pivot_longer(cols = c(a_med, c_med, d_med),
                      names_to = "Parameter", values_to = "Median") |>
  mutate(Parameter = sub("_med$", "", Parameter)) |>
  group_by(Model, Parameter) |>
  summarise(
    BMD10 = Median[BMR == "BMD10"],
    BMD50 = Median[BMR == "BMD50"],
    `Shift (%)` = round(100 * abs(BMD50 - BMD10) / abs(BMD10), 2),
    .groups = "drop"
  ) |>
  dplyr::select(Model, Parameter, BMD10, BMD50, `Shift (%)`) |>
  arrange(Parameter, desc(`Shift (%)`))

ktbl(shift_summary,
     caption = "Shift in 'invariant' parameters between BMD10 and BMD50 parameterisations")
```

Non-zero values in the `Shift (%)` column confirm that the BMR
parameterisation affects **individual model estimates**, not only the model
weights. The combined effect on the BMA result is the product of both:
different weights applied to different curves.

The data confirm it clearly: the BMR parameterisation shifts **all** curve-shape parameters, not just the weights.

Key takeaways from above tables:

**Parameters that should be invariant but aren't:**

- **`c` (fold change)** shows the largest shifts — up to **53%** for IE4_N and **46%** for LN4_N. This is the asymptote ratio, a property of the curve that should not depend on which BMR level you use to parameterise.

- **`d` (Hill power)** shifts 21–39% across all models. This controls the steepness of the curve — again, a curve property, not a BMR property.

- **`a` (background)** is relatively stable for the first 6 models (< 2.5%), but **P4_N and L4_N** show 28–46% shifts. These two models have a fundamentally different parameterisation (negative `c`, large `a`), and the reparameterisation hits them hardest.

**Why this happens:**

The uniform prior on `par2 = BMD_q` implies a **different prior on `b`** for each `q`, because $b = f(\text{BMD}_q, c, d, q)$. Since `b`, `c`, and `d` are correlated in the posterior (they jointly determine the curve shape), a shift in the effective prior on `b` propagates into the marginals of `c` and `d`. The posterior is not dominated enough by the likelihood (only 5 dose groups, ~10 animals each) to override these prior-induced shifts.

**The bottom line for EFSA argument:**

The effect of BMR parameterisation on the final BMA result operates through **two channels**:

1. **Different model weights** (already shown in the weights table)
2. **Different parameter estimates within each model** (now quantified — shifts up to 53%)

Both are consequences of the same root cause: `q` enters the prior implicitly through the reparameterisation, making each run a genuinely different Bayesian analysis rather than just a different summary of the same posterior. This strengthens the case for either (a) parameterising at BMD50 for stability, or (b) moving to a `(a, b, c, d)` parameterisation where `q` only enters at the post-processing stage.

## Model-averaged predicted response at BMD estimates

For each BMR level, we extract the model-averaged predicted response at the
corresponding BMD estimate. This uses `predict.BMADR()` with
`what = "resp_at_BMD"`. We compare Bridge Sampling (from `sampling_MA()`)
with Laplace approximation (from `full.laplace_MA()`).

```{r resp-at-bmd}
# Bridge Sampling: predict response at BMD for each BMR level
resp_bs10 <- predict.BMADR(fit10, type = "increasing", what = "resp_at_BMD",
                            model_averaged = TRUE, weight_type = "BS")
resp_bs20 <- predict.BMADR(fit20, type = "increasing", what = "resp_at_BMD",
                            model_averaged = TRUE, weight_type = "BS")
resp_bs50 <- predict.BMADR(fit50, type = "increasing", what = "resp_at_BMD",
                            model_averaged = TRUE, weight_type = "BS")

# Laplace: predict response at BMD for each BMR level
resp_lp10 <- predict.BMADR(lp10, type = "increasing", what = "resp_at_BMD",
                            model_averaged = TRUE, weight_type = "LP")
resp_lp20 <- predict.BMADR(lp20, type = "increasing", what = "resp_at_BMD",
                            model_averaged = TRUE, weight_type = "LP")
resp_lp50 <- predict.BMADR(lp50, type = "increasing", what = "resp_at_BMD",
                            model_averaged = TRUE, weight_type = "LP")

# Combine into a comparison table
resp_table <- bind_rows(
  tibble(BMR = "10%", Method = "Bridge Sampling",
         BMD = resp_bs10$model_averaged$BMD,
         `Response at BMD` = resp_bs10$model_averaged$model_averaged_response),
  tibble(BMR = "20%", Method = "Bridge Sampling",
         BMD = resp_bs20$model_averaged$BMD,
         `Response at BMD` = resp_bs20$model_averaged$model_averaged_response),
  tibble(BMR = "50%", Method = "Bridge Sampling",
         BMD = resp_bs50$model_averaged$BMD,
         `Response at BMD` = resp_bs50$model_averaged$model_averaged_response),
  tibble(BMR = "10%", Method = "Laplace",
         BMD = resp_lp10$model_averaged$BMD,
         `Response at BMD` = resp_lp10$model_averaged$model_averaged_response),
  tibble(BMR = "20%", Method = "Laplace",
         BMD = resp_lp20$model_averaged$BMD,
         `Response at BMD` = resp_lp20$model_averaged$model_averaged_response),
  tibble(BMR = "50%", Method = "Laplace",
         BMD = resp_lp50$model_averaged$BMD,
         `Response at BMD` = resp_lp50$model_averaged$model_averaged_response)
) |>
  mutate(across(c(BMD, `Response at BMD`), ~ round(.x, 4)))

resp_table |>
  tidyr::pivot_wider(
    names_from  = Method,
    values_from = c(BMD, `Response at BMD`),
    names_glue  = "{Method}_{.value}"
  )
```

## Model-averaged dose-response: Laplace vs Bridge Sampling

`predict.BMADR()` with `what = "predicted"` computes the dose-response at a
fine dose grid using median posterior parameters per model, then blends the
16 curves with model weights. We compare the two weighting approaches.

```{r ma-prediction, fig.width=9, fig.height=6}
# predict.BMADR expects doses on the RESCALED scale (dose / maxDose)
maxD <- data_N10$data$maxD
pseudo_zero <- 0.01
dose_grid_rescaled <- seq(pseudo_zero / maxD, 1, length.out = 200)

# Bridge Sampling model-averaged curve (from sampling_MA fit)
pred_bs <- predict.BMADR(
  fit10, dose = dose_grid_rescaled, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "BS"
)

# Laplace model-averaged curve (from full.laplace_MA fit)
pred_lp <- predict.BMADR(
  lp10, dose = dose_grid_rescaled, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "LP"
)

# Convert rescaled Dose back to original scale for plotting
pred_bs$predicted$Dose     <- pred_bs$predicted$Dose * maxD
pred_bs$model_averaged$Dose <- pred_bs$model_averaged$Dose * maxD
pred_lp$predicted$Dose     <- pred_lp$predicted$Dose * maxD
pred_lp$model_averaged$Dose <- pred_lp$model_averaged$Dose * maxD

# Combine curves
ma_curves <- bind_rows(
  pred_bs$model_averaged |> mutate(Method = "Bridge Sampling"),
  pred_lp$model_averaged |> mutate(Method = "Laplace")
)

# BMD10 estimates from both methods (already on original scale)
bmd10_bs  <- fit10$MA_bridge_sampling[["BMD"]]
bmd10_lp  <- lp10$MA[2]

bmd_lines <- tibble(
  Method = c("Bridge Sampling", "Laplace"),
  BMD    = c(bmd10_bs, bmd10_lp)
)

# Shift dose = 0 for log-scale display (pseudo_zero defined above)
dat_plot <- dplyr::mutate(dat, Dose_plot = ifelse(Dose == 0, pseudo_zero, Dose))

ggplot() +
  # Individual model curves from BS fit (background context)
  geom_line(
    data = pred_bs$predicted,
    aes(x = Dose, y = predicted, group = Model),
    colour = "grey80", linewidth = 0.3
  ) +
  # Model-averaged curves
  geom_line(
    data = ma_curves,
    aes(x = Dose, y = model_averaged, colour = Method),
    linewidth = 1.2
  ) +
  # Observed data
  geom_point(
    data = dat_plot,
    aes(x = Dose_plot, y = Mean),
    size = 3
  ) +
  geom_errorbar(
    data = dat_plot,
    aes(x = Dose_plot,
        ymin = Mean - 1.96 * SD / sqrt(n),
        ymax = Mean + 1.96 * SD / sqrt(n)),
    width = 0.05
  ) +
  # BMD10 vertical lines
  geom_vline(
    data = bmd_lines,
    aes(xintercept = BMD, colour = Method),
    linetype = "dashed", linewidth = 0.7
  ) +
  scale_colour_manual(
    values = c("Bridge Sampling" = "steelblue", "Laplace" = "darkorange")
  ) +
  scale_x_log10() +
  labs(
    title    = "Model-averaged dose-response: Bridge Sampling vs Laplace (BMD10)",
    subtitle = "Grey: 16 individual model curves (BS median parameters);\ndashed lines: BMD10 estimates. Control (dose = 0) plotted at pseudo-zero.",
    x        = "Dose (log scale)", y = "Mean response",
    colour   = "Method"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Side-by-side comparison across all BMR levels

```{r ma-prediction-facet, fig.width=10, fig.height=5}
# Helper: predict on rescaled grid, then convert Dose back to original scale
predict_ma <- function(mod_obj, dose_rescaled, maxD, method, bmr, wt) {
  pred <- predict.BMADR(
    mod_obj, dose = dose_rescaled, type = "increasing",
    what = "predicted", model_averaged = TRUE, weight_type = wt
  )
  pred$model_averaged$Dose <- pred$model_averaged$Dose * maxD
  pred$model_averaged |> mutate(BMR = bmr, Method = method)
}

pred_all <- bind_rows(
  predict_ma(fit10, dose_grid_rescaled, maxD, "Bridge Sampling", "10%", "BS"),
  predict_ma(fit20, dose_grid_rescaled, data_N20$data$maxD, "Bridge Sampling", "20%", "BS"),
  predict_ma(fit50, dose_grid_rescaled, data_N50$data$maxD, "Bridge Sampling", "50%", "BS"),
  predict_ma(lp10,  dose_grid_rescaled, maxD, "Laplace", "10%", "LP"),
  predict_ma(lp20,  dose_grid_rescaled, data_N20$data$maxD, "Laplace", "20%", "LP"),
  predict_ma(lp50,  dose_grid_rescaled, data_N50$data$maxD, "Laplace", "50%", "LP")
)

# BMD vertical lines for all combinations
bmd_vlines <- bind_rows(bmd_all) |>
    dplyr::select(BMR, Method, BMD)

ggplot() +
  geom_line(
    data = pred_all,
    aes(x = Dose, y = model_averaged, colour = Method),
    linewidth = 1
  ) +
  geom_point(
    data = dat_plot,
    aes(x = Dose_plot, y = Mean),
    size = 2.5
  ) +
  geom_errorbar(
    data = dat_plot,
    aes(x = Dose_plot,
        ymin = Mean - 1.96 * SD / sqrt(n),
        ymax = Mean + 1.96 * SD / sqrt(n)),
    width = 0.05
  ) +
  geom_vline(
    data = bmd_vlines,
    aes(xintercept = BMD, colour = Method),
    linetype = "dashed", linewidth = 0.6
  ) +
  facet_wrap(~BMR) +
  scale_colour_manual(
    values = c("Bridge Sampling" = "steelblue", "Laplace" = "darkorange")
  ) +
  scale_x_log10() +
  labs(
    title    = "Model-averaged dose-response curves by BMR level and method",
    subtitle = "Dashed lines: BMD estimates; points: observed data ± 95% CI",
    x        = "Dose (log scale)", y = "Mean response",
    colour   = "Method"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Goodness of fit: Bayes factor vs saturated model

BMABMDR tests goodness of fit by comparing the best-fitting dose-response
model against a saturated (ANOVA) model via a BIC-based Bayes factor.

**Note on direction:** Inspection of `modelTest()` in the package source
reveals that the internally computed BF is
$\text{BF} = \exp\!\bigl(-\tfrac{1}{2}(\text{BIC}_{\text{DR}} - \text{BIC}_{\text{sat}})\bigr)$,
which is BF in favour of the **DR model** (BF > 1 ⇒ DR preferred). However,
the returned `$bf` value (and the printed message) report **1/BF**, i.e. the
Bayes factor in favour of the **saturated model**. Therefore a value < 1
means the DR model is preferred. The package flags inadequate fit only when
`$bf > 10` (saturated model preferred 10 : 1).

```{r gof-bf}
gof_table <- tibble(
  BMR    = rep(c("10%", "20%", "50%"), 2),
  Method = rep(c("Bridge Sampling", "Laplace"), each = 3),
  `BF (sat / DR)` = c(
    fit10$bf, fit20$bf, fit50$bf,
    lp10$bf,  lp20$bf,  lp50$bf
  ),
  `BF (DR / sat)` = round(1 / `BF (sat / DR)`, 3),
  Interpretation = ifelse(`BF (sat / DR)` < 1,
                          "DR model preferred",
                          "Saturated marginally preferred"),
  `GOF message` = c(
    fit10$gof_check, fit20$gof_check, fit50$gof_check,
    lp10$gof_check,  lp20$gof_check,  lp50$gof_check
  )
) |>
  mutate(`BF (sat / DR)` = round(`BF (sat / DR)`, 4))

ktbl(gof_table, caption = "Goodness of fit: BF as returned by BMABMDR (BF = saturated / DR)")
```

For Bridge Sampling, `BF(sat/DR) < 1` across all BMR levels, meaning the
dose-response model is preferred over the saturated model — the fit is
adequate. For Laplace, `BF(sat/DR) ≈ 1.09`, meaning the two models are
roughly equally supported; this is within the "fits sufficiently well"
range. The BS fit thus provides slightly better GOF by this criterion.

## Per-model log-likelihoods

```{r gof-loglik}
model_names <- c("E4_N", "IE4_N", "H4_N", "LN4_N",
                 "G4_N", "QE4_N", "P4_N", "L4_N")

ll_table <- tibble(
  Model = model_names,
  `BS BMD10 (llN)` = round(fit10$llN, 2),
  `BS BMD20 (llN)` = round(fit20$llN, 2),
  `BS BMD50 (llN)` = round(fit50$llN, 2),
  `LP BMD10 (llN)` = round(lp10$llN, 2),
  `LP BMD20 (llN)` = round(lp20$llN, 2),
  `LP BMD50 (llN)` = round(lp50$llN, 2)
)

ktbl(ll_table, caption = "Per-model log-likelihoods (Normal)")
```

## Residual fit: weighted RSS at observed doses

To quantify how well the model-averaged curve matches the data, we compute
the weighted residual sum of squares (RSS) at the observed dose levels for
each BMR parameterisation × method combination. Weights are $n_i / \text{SD}_i^2$
(inverse variance of the group mean).

```{r residual-rss}
# Predict at observed (non-zero) rescaled doses for each fit
obs_doses <- dat$Dose
obs_rescaled <- obs_doses / maxD
# For dose = 0 use a very small positive value (models are defined at 0)
obs_rescaled[obs_rescaled == 0] <- 1e-6

compute_rss <- function(mod_obj, dose_rescaled, maxD_fit, obs_mean, obs_sd, obs_n, wt) {
  pred <- predict.BMADR(
    mod_obj, dose = dose_rescaled, type = "increasing",
    what = "predicted", model_averaged = TRUE, weight_type = wt
  )
  fitted_vals <- pred$model_averaged$model_averaged
  w <- obs_n / obs_sd^2
  sum(w * (obs_mean - fitted_vals)^2)
}

rss_results <- tibble(
  Parameterisation = rep(c("BMD10", "BMD20", "BMD50"), 2),
  Method = rep(c("Bridge Sampling", "Laplace"), each = 3),
  `Weighted RSS` = round(c(
    compute_rss(fit10, obs_rescaled, maxD, dat$Mean, dat$SD, dat$n, "BS"),
    compute_rss(fit20, obs_rescaled, data_N20$data$maxD, dat$Mean, dat$SD, dat$n, "BS"),
    compute_rss(fit50, obs_rescaled, data_N50$data$maxD, dat$Mean, dat$SD, dat$n, "BS"),
    compute_rss(lp10,  obs_rescaled, maxD, dat$Mean, dat$SD, dat$n, "LP"),
    compute_rss(lp20,  obs_rescaled, data_N20$data$maxD, dat$Mean, dat$SD, dat$n, "LP"),
    compute_rss(lp50,  obs_rescaled, data_N50$data$maxD, dat$Mean, dat$SD, dat$n, "LP")
  ), 3)
)

rss_results |>
  tidyr::pivot_wider(names_from = Method, values_from = `Weighted RSS`) |>
  ktbl(caption = "Weighted RSS at observed doses")
```

Lower weighted RSS indicates a better match between the model-averaged curve
and the observed data at the tested dose levels. Differences across BMR
parameterisations reflect the `q`-dependent model weights, not inherently
better model fit — the data and likelihood are identical in all runs.

## BMD10 estimates: detailed comparison

BMD10 estimates from both methods, model-averaged predicted response at the
BMD10 dose, and percentage increase over the control group mean.

```{r bmd10-detail-table}
# Control mean: average of the two dose-0 groups
control_mean <- mean(dat$Mean[dat$Dose == 0])

# Model-averaged predicted background (response at dose ≈ 0)
bg_rescaled <- 1e-6  # near-zero on rescaled scale
bg_bs <- predict.BMADR(
  fit10, dose = bg_rescaled, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "BS"
)$model_averaged$model_averaged
bg_lp <- predict.BMADR(
  lp10, dose = bg_rescaled, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "LP"
)$model_averaged$model_averaged

# Predicted response at BMD10 from both methods
resp_at_bmd10_bs <- resp_bs10$model_averaged$model_averaged_response
resp_at_bmd10_lp <- resp_lp10$model_averaged$model_averaged_response

tibble(
  Method           = c("Bridge Sampling", "Laplace"),
  BMDL             = c(bmd_bs$BMDL[1], bmd_lp$BMDL[1]),
  BMD10            = c(bmd_bs$BMD[1],  bmd_lp$BMD[1]),
  BMDU             = c(bmd_bs$BMDU[1], bmd_lp$BMDU[1]),
  `Response at BMD10` = round(c(resp_at_bmd10_bs, resp_at_bmd10_lp), 4),
  `Model bg`       = round(c(bg_bs, bg_lp), 4),
  `Obs control`    = round(control_mean, 4),
  `% vs model bg`  = round(100 * (c(resp_at_bmd10_bs, resp_at_bmd10_lp) -
                                    c(bg_bs, bg_lp)) / c(bg_bs, bg_lp), 2),
  `% vs obs control` = round(100 * (c(resp_at_bmd10_bs, resp_at_bmd10_lp) -
                                      control_mean) / control_mean, 2)
) |>
  ktbl(caption = "BMD10 detailed comparison: Bridge Sampling vs Laplace", digits = 4)
```

## Cross-BMR prediction: response at BMD10 from different parameterisations

We take the BMD10 estimate from the Bridge Sampling / BMD10 parameterisation
and evaluate the model-averaged dose-response from the BMD20 and BMD50
parameterised fits at that same dose. This tests whether the predicted
response at a given dose is consistent across parameterisations.

```{r cross-bmr-prediction}
# BMD10 from the BS/BMD10 fit (original scale)
bmd10_dose <- fit10$MA_bridge_sampling[["BMD"]]

# Rescale to [0,1] for each fit's maxD
bmd10_rescaled_10 <- bmd10_dose / data_N10$data$maxD
bmd10_rescaled_20 <- bmd10_dose / data_N20$data$maxD
bmd10_rescaled_50 <- bmd10_dose / data_N50$data$maxD

# Predict at that single dose using each fit
pred_at_bmd10_bs10 <- predict.BMADR(
  fit10, dose = bmd10_rescaled_10, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "BS"
)$model_averaged$model_averaged

pred_at_bmd10_bs20 <- predict.BMADR(
  fit20, dose = bmd10_rescaled_20, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "BS"
)$model_averaged$model_averaged

pred_at_bmd10_bs50 <- predict.BMADR(
  fit50, dose = bmd10_rescaled_50, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "BS"
)$model_averaged$model_averaged

pred_at_bmd10_lp10 <- predict.BMADR(
  lp10, dose = bmd10_rescaled_10, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "LP"
)$model_averaged$model_averaged

pred_at_bmd10_lp20 <- predict.BMADR(
  lp20, dose = bmd10_rescaled_20, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "LP"
)$model_averaged$model_averaged

pred_at_bmd10_lp50 <- predict.BMADR(
  lp50, dose = bmd10_rescaled_50, type = "increasing",
  what = "predicted", model_averaged = TRUE, weight_type = "LP"
)$model_averaged$model_averaged

tibble(
  `Parameterisation` = rep(c("BMD10", "BMD20", "BMD50"), 2),
  Method             = rep(c("Bridge Sampling", "Laplace"), each = 3),
  `Dose evaluated`   = round(bmd10_dose, 3),
  `Predicted response` = round(c(pred_at_bmd10_bs10, pred_at_bmd10_bs20,
                                  pred_at_bmd10_bs50,
                                  pred_at_bmd10_lp10, pred_at_bmd10_lp20,
                                  pred_at_bmd10_lp50), 4),
  `% vs control`     = round(100 * (c(pred_at_bmd10_bs10, pred_at_bmd10_bs20,
                                       pred_at_bmd10_bs50,
                                       pred_at_bmd10_lp10, pred_at_bmd10_lp20,
                                       pred_at_bmd10_lp50) - control_mean) / control_mean, 2)
) |>
  tidyr::pivot_wider(
    names_from = Method,
    values_from = c(`Predicted response`, `% vs control`),
    names_glue = "{Method}_{.value}"
  ) |>
  ktbl(caption = "Cross-BMR prediction: response at BMD10 dose from different parameterisations")
```

If the parameterisation were irrelevant, all rows would give identical
predictions at the same dose. Differences reflect the `q`-specific model
weights and the different effective priors.

## Posterior stability across BMR parameterisations

A common understanding in dose-response modelling is that parameterising at
a higher BMR (e.g. 50%) yields more stable inference because the BMD lies
closer to the inflection region of the curve, where the data carry more
information. We can quantify this by comparing the coefficient of variation
(CV) of the model-averaged BMD posterior across the three parameterisations.

```{r bmd-stability}
stability_table <- tibble(
  BMR = c("10%", "20%", "50%"),
  `Posterior mean` = c(
    mean(fit10$BMDMixtureBS),
    mean(fit20$BMDMixtureBS),
    mean(fit50$BMDMixtureBS)
  ),
  `Posterior SD` = c(
    sd(fit10$BMDMixtureBS),
    sd(fit20$BMDMixtureBS),
    sd(fit50$BMDMixtureBS)
  ),
  `CV (%)` = round(100 * `Posterior SD` / `Posterior mean`, 1),
  `Weighted RSS (BS)` = c(
    compute_rss(fit10, obs_rescaled, maxD, dat$Mean, dat$SD, dat$n, "BS"),
    compute_rss(fit20, obs_rescaled, data_N20$data$maxD, dat$Mean, dat$SD, dat$n, "BS"),
    compute_rss(fit50, obs_rescaled, data_N50$data$maxD, dat$Mean, dat$SD, dat$n, "BS")
  ),
  `Weighted RSS (LP)` = c(
    compute_rss(lp10, obs_rescaled, maxD, dat$Mean, dat$SD, dat$n, "LP"),
    compute_rss(lp20, obs_rescaled, data_N20$data$maxD, dat$Mean, dat$SD, dat$n, "LP"),
    compute_rss(lp50, obs_rescaled, data_N50$data$maxD, dat$Mean, dat$SD, dat$n, "LP")
  )
) |>
  mutate(across(c(`Posterior mean`, `Posterior SD`, `Weighted RSS (BS)`,
                   `Weighted RSS (LP)`), ~ round(.x, 3)))

ktbl(stability_table,
     caption = "Posterior stability of model-averaged BMD across parameterisations")
```

A lower CV indicates the BMD is better identified by the data under that
parameterisation. Intuitively, at BMD50 the dose is closer to the inflection
point of the dose-response curve where the gradient is steepest and the data
carry the most information about the curve shape. This makes the BMD parameter
better constrained by the likelihood, reducing its dependence on the prior.

The weighted RSS column provides a complementary view: a lower RSS for the
BMD50 parameterisation indicates that the resulting model-averaged curve also
fits the observed data slightly better, consistent with the weights being
more data-driven and less prior-influenced at higher BMR levels.

---

# Frequentist comparison: `drc` and `bmd` packages

## Best model fit with `drc`

We use the `drc` package to fit several standard dose-response models to
the same data (using the summary statistics to construct a weighted
regression).

```{r drc-fit}
# drc works with individual-level data or weighted means.
# We use the summary data with weights = n per group.
dat_drc <- dat |>
  mutate(
    Var   = SD^2 / n,  # variance of the mean
    lower = Mean - 1.96 * SD / sqrt(n),
    upper = Mean + 1.96 * SD / sqrt(n)
  )

# Fit several candidate models
fit_ll4 <- drm(Mean ~ Dose, data = dat_drc, weights = n,
               fct = LL.4(), type = "continuous")
fit_ll3 <- drm(Mean ~ Dose, data = dat_drc, weights = n,
               fct = LL.3(), type = "continuous")
fit_w14 <- drm(Mean ~ Dose, data = dat_drc, weights = n,
               fct = W1.4(), type = "continuous")
fit_w23 <- drm(Mean ~ Dose, data = dat_drc, weights = n,
               fct = W2.3(), type = "continuous")
fit_w24 <- try(drm(Mean ~ Dose, data = dat_drc, weights = n,
                    fct = W2.4(), type = "continuous"), silent = TRUE)
fit_exp <- try(drm(Mean ~ Dose, data = dat_drc, weights = n,
                   fct = EXD.3(), type = "continuous"), silent = TRUE)

# Model selection by AIC
models_drc <- list(LL.4 = fit_ll4, LL.3 = fit_ll3, W1.4 = fit_w14, W2.3 = fit_w23)
if (!inherits(fit_w24, "try-error")) models_drc$W2.4 <- fit_w24
if (!inherits(fit_exp, "try-error")) models_drc$EXD.3 <- fit_exp

aic_drc <- tibble(
  Model = names(models_drc),
  AIC   = sapply(models_drc, AIC),
  logLik = sapply(models_drc, logLik)
) |> arrange(AIC)

ktbl(aic_drc, caption = "drc model comparison by AIC", digits = 2)
```

```{r drc-best-plot, fig.width=9, fig.height=6}
best_model_name <- aic_drc$Model[1]
best_fit <- models_drc[[best_model_name]]

# drc prediction on fine dose grid (original scale)
drc_grid <- data.frame(Dose = seq(0.01, max(dat$Dose), length.out = 200))
drc_grid$predicted <- predict(best_fit, newdata = drc_grid)

# BMD10 for the best drc model (computed here so it's available for the plot)
bmd_best_drc_obj <- try(
  bmd(best_fit, bmr = 0.10, backgType = "modelBased",
      def = "relative", interval = "inv"),
  silent = TRUE
)
bmd_best_drc <- if (!inherits(bmd_best_drc_obj, "try-error")) {
  bmd_best_drc_obj$Results[1]
} else NA_real_

# Combine drc + Bayesian MA curves
all_curves <- bind_rows(
  drc_grid |> mutate(Method = paste0("drc best (", best_model_name, ")")),
  ma_curves |> rename(predicted = model_averaged)
)

# BMD10 vertical lines — Bayesian + drc best model
bmd_vlines_all <- tibble(
  Method = c("Bridge Sampling", "Laplace",
             paste0("drc best (", best_model_name, ")")),
  BMD    = c(bmd10_bs, bmd10_lp, bmd_best_drc)
) |> filter(!is.na(BMD))

ggplot() +
  # Dose-response curves
  geom_line(
    data = all_curves,
    aes(x = Dose, y = predicted, colour = Method),
    linewidth = 1.1
  ) +
  # Observed data
  geom_point(
    data = dat_plot,
    aes(x = Dose_plot, y = Mean),
    size = 3
  ) +
  geom_errorbar(
    data = dat_plot,
    aes(x = Dose_plot,
        ymin = Mean - 1.96 * SD / sqrt(n),
        ymax = Mean + 1.96 * SD / sqrt(n)),
    width = 0.05
  ) +
  # BMD10 vertical lines
  geom_vline(
    data = bmd_vlines_all,
    aes(xintercept = BMD, colour = Method),
    linetype = "dashed", linewidth = 0.7
  ) +
  scale_colour_manual(values = c(
    "Bridge Sampling" = "steelblue",
    "Laplace"         = "darkorange",
    setNames("forestgreen", paste0("drc best (", best_model_name, ")"))
  )) +
  scale_x_log10() +
  labs(
    title    = "Bayesian model-averaged vs drc best-fit dose-response (BMD10)",
    subtitle = paste0("Dashed lines: BMD10 estimates. drc best model: ",
                      best_model_name, " (AIC = ", round(aic_drc$AIC[1], 2), ")"),
    x = "Dose (log scale)", y = "Mean response",
    colour = "Method"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## BMD estimation with `bmd` package (frequentist model averaging)

The `bmd` package provides frequentist BMD estimation with model averaging
using the `bmdMA` function or single-model BMD via `bmd()`. We use inverse
regression (`interval = "inv"`) for BMDL confidence intervals, and fit all
four-parameter models plus EXD.3 (LL.4, W1.4, W2.4, and EXD.3).

```{r bmd-estimation}
# Use all fitted models for per-model BMD estimation
good_models <- list(LL.4 = fit_ll4, W1.4 = fit_w14)
if (!inherits(fit_w24, "try-error")) good_models$W2.4 <- fit_w24
if (!inherits(fit_exp, "try-error")) good_models$EXD.3 <- fit_exp

# bmd() with inverse regression for BMDL
bmd_results <- lapply(good_models, function(m) {
  try(bmd(m, bmr = 0.10, backgType = "modelBased",
          def = "relative", interval = "inv"), silent = TRUE)
})
bmd_results <- bmd_results[!sapply(bmd_results, inherits, "try-error")]

bmd_freq <- tibble(
  Model  = names(bmd_results),
  BMD    = as.numeric(sapply(bmd_results, function(x) x$Results[1])),
  BMDL   = as.numeric(sapply(bmd_results, function(x) x$Results[2]))
) |>
  mutate(
    # Model-predicted background (response at dose = 0)
    `Model bg` = round(mapply(
      function(nm) predict(good_models[[nm]],
                           newdata = data.frame(Dose = 0)),
      Model
    ), 4),
    # Predicted response at BMD from each model
    `Response at BMD` = round(mapply(
      function(nm, bmd_val) predict(good_models[[nm]],
                                    newdata = data.frame(Dose = bmd_val)),
      Model, BMD
    ), 4),
    `% vs model bg` = round(100 * (`Response at BMD` - `Model bg`) / `Model bg`, 2),
    `% vs obs control` = round(100 * (`Response at BMD` - control_mean) /
                                      control_mean, 2),
    across(c(BMD, BMDL), ~ round(.x, 3))
  )

ktbl(bmd_freq, caption = "Frequentist per-model BMD10 estimates (bmd package)")
```

```{r bmd-ma}
# Model-averaged BMD using bmd::bmdMA with AIC weights
# type = "Kang": weighted average of per-model BMD and BMDL
# type = "Buckland": requires SE (delta method), which produces NaN for this dataset
# interval = "inv" for inverse regression BMDL (consistent with per-model estimates above)
bmd_ma_result <- try(
  bmdMA(modelList = unname(good_models),
        modelWeights = "AIC",
        bmr = 0.10, backgType = "modelBased",
        def = "relative", interval = "inv",
        type = "Kang"),
  silent = TRUE
)

if (!inherits(bmd_ma_result, "try-error") &&
    !any(is.nan(bmd_ma_result$Results))) {
  tibble(
    Method = "bmd MA (Kang, AIC weights)",
    BMD    = round(bmd_ma_result$Results[1], 3),
    BMDL   = round(bmd_ma_result$Results[2], 3)
  ) |>
    ktbl(caption = "Frequentist model-averaged BMD10 (bmdMA, Kang method, AIC weights)")
} else {
  cat("bmdMA() returned an error:\n", as.character(bmd_ma_result))
}
```



A quick manual check right after the `bmd-ma` chunk. The Kang method is just a weighted average of per-model BMD and BMDL values.

```{r}
# Manual AIC-weighted BMD (Kang method = weighted average)
aic_vals <- sapply(good_models, AIC)
aic_wts <- exp(-(aic_vals - min(aic_vals))) / sum(exp(-(aic_vals - min(aic_vals))))

# Per-model BMD and BMDL from bmd_results (already computed above)
manual_bmd  <- sum(aic_wts[names(bmd_results)] * sapply(bmd_results, \(x) x$Results[1]))
manual_bmdl <- sum(aic_wts[names(bmd_results)] * sapply(bmd_results, \(x) x$Results[2]))

tibble(
  Model  = names(aic_wts),
  `AIC weight` = round(aic_wts, 6),
  BMD    = round(sapply(good_models, \(m) {
    r <- bmd_results[[names(which(sapply(good_models, identical, m)))]]; 
    if (!is.null(r)) r$Results[1] else NA
  }), 3),
  BMDL   = round(sapply(good_models, \(m) {
    r <- bmd_results[[names(which(sapply(good_models, identical, m)))]]; 
    if (!is.null(r)) r$Results[2] else NA
  }), 3)
) |> ktbl(caption = "AIC weights and per-model estimates")

# Weighted averages
tibble(
  `Manual BMD_MA`  = round(manual_bmd, 3),
  `Manual BMDL_MA` = round(manual_bmdl, 3),
  `bmdMA BMD`      = round(bmd_ma_result$Results[1], 3),
  `bmdMA BMDL`     = round(bmd_ma_result$Results[2], 3)
) |> ktbl(caption = "Manual vs bmdMA check")
```

This computes AIC weights the standard way ($w_i = e^{-\Delta_i/2} / \sum e^{-\Delta_j/2}$) and takes the weighted average of per-model BMD/BMDL — exactly what Kang does. The two rows should match.

## Summary comparison: Bayesian vs frequentist BMD10

```{r final-comparison}
summary_comparison <- bind_rows(
  tibble(
    Approach = "BMABMDR (Bridge Sampling)",
    BMDL     = round(bmd_bs$BMDL[1], 3),
    BMD      = round(bmd_bs$BMD[1], 3),
    BMDU     = round(bmd_bs$BMDU[1], 3)
  ),
  tibble(
    Approach = "BMABMDR (Laplace)",
    BMDL     = round(bmd_lp$BMDL[1], 3),
    BMD      = round(bmd_lp$BMD[1], 3),
    BMDU     = round(bmd_lp$BMDU[1], 3)
  ),
  if (!inherits(bmd_ma_result, "try-error") &&
      !any(is.nan(bmd_ma_result$Results))) {
    tibble(
      Approach = "bmd (Kang MA, AIC weights)",
      BMDL     = round(bmd_ma_result$Results[2], 3),
      BMD      = round(bmd_ma_result$Results[1], 3),
      BMDU     = NA_real_
    )
  },
  tibble(
    Approach = paste0("drc best model (", best_model_name, ")"),
    BMDL     = if (exists("bmd_results") && best_model_name %in% names(bmd_results))
                 round(bmd_results[[best_model_name]]$Results[2], 3) else NA_real_,
    BMD      = if (exists("bmd_results") && best_model_name %in% names(bmd_results))
                 round(bmd_results[[best_model_name]]$Results[1], 3) else NA_real_,
    BMDU     = NA_real_
  )
)

ktbl(summary_comparison, caption = "Summary: Bayesian vs frequentist BMD10 estimates")
```

---

# Summary of findings

| Issue | Detail |
|---|---|
| **One run per BMR** | `q` is fixed data in Stan; 16 models refitted per BMR level |
| **Different averaged models** | Prior hyperparameters and hence model weights are `q`-specific |
| **Prior family** | PERT on rescaled BMD; default shape $g = 0.0001$ ≈ Uniform |
| **Support** | $(0,\, 3 \cdot \max(\text{dose}))$ — upper bound depends on data |
| **Non-informative claim** | Approximately flat shape, but support is data-adaptive and `q`-specific; not invariant to log-scale reparameterisation |
| **Post-hoc inversion** | Feasible with HMC draws; requires per-model analytical inversion; uses weights from the original `q` run (approximate) |
| **Laplace vs Bridge Sampling** | Both approaches yield similar BMD estimates; BS provides full posterior draws for post-hoc analyses |
| **Cross-BMR consistency** | Predicted response at BMD10 is similar across BMD10/BMD20/BMD50 parameterisations, but not identical |
| **Frequentist comparison** | `drc`/`bmd` package BMD10 estimates provide an independent benchmark for the Bayesian model-averaged results |
| **`predict.BMADR` scale** | Dose argument must be on **rescaled** scale (÷ maxDose); multiply back for plotting — undocumented |

## Regulatory relevance

Current EFSA and EPA guidance specifies a **single pre-specified BMR** (10%
for continuous endpoints), so the one-run-per-BMR design poses no practical
problem for standard regulatory submissions. Multiple BMR sensitivity analysis
is exploratory, and the differences in model weights across BMR levels are
typically small when data are informative. However, for transparent reporting
the distinction between "three separate BMA analyses" and "one BMA analysis
evaluated at three BMR levels" should be made explicit.

An alternative design — parameterising Stan in terms of `(a, b, c, d)` and
computing any `BMD_q` as a `generated quantities` post-processing step — would
give a single coherent joint posterior from which all BMR levels could be
derived without refitting. This would eliminate both the multiple-run
requirement and the `q`-dependence of the prior. It would, however, require
rethinking prior elicitation, since a prior on `b` is less interpretable than
a prior on `BMD_q` directly.

---

# Practical implications for the EFSA BMD online tool

## Choosing between Laplace and Bridge Sampling

The EFSA online tool offers both Laplace approximation and full HMC (Bridge
Sampling) inference. In routine use, the Laplace approximation is faster, more
robust, and in our experience more often produces sensible results — the HMC
sampler can struggle with multimodal posteriors, poor initialisation, or
near-boundary parameters, leading to divergent transitions or poor mixing
that inflate posterior uncertainty. In the present example the CV of the
Bridge Sampling posterior at BMD10 exceeds 180%, indicating a heavy tail
driven by poorly identified models; Laplace, which relies on a Gaussian
approximation around the MAP, is less affected by such tails.

That said, the present example is a case where Bridge Sampling arguably gives
a more defensible result: it provides the full model-averaged posterior
distribution, including asymmetry and multimodality, whereas Laplace can only
report point estimates and symmetric intervals. When both methods converge
well and produce comparable BMD estimates, either is acceptable. When they
disagree substantially, the practitioner should investigate which models are
driving the difference and whether the HMC diagnostics (divergences, $\hat{R}$,
effective sample size) indicate reliable sampling.

**Recommendation.** Use the Laplace approximation as the default for
regulatory submissions; switch to Bridge Sampling when a full posterior is
needed (e.g. for post-hoc BMR inversion) or when the Laplace result appears
unstable. Always compare both when feasible.

## BMD50 parameterisation gives more stable estimation

The stability analysis in this document shows that the BMD50
parameterisation yields a dramatically lower CV (65% vs 189% for BMD10) and
a lower weighted RSS, indicating that the model-averaged curve is better
constrained by the data. This is expected: at BMR 50%, the BMD lies in the
steepest region of the dose-response curve where the data carry the most
information, making the posterior less sensitive to the prior and to the
choice of inference method.

Critically, the choice of parameterisation also affects the **BMD10 estimate
itself**. As shown in the cross-BMR prediction table, the model-averaged
response at the BMD10 dose varies by up to 6 percentage points depending on
whether the underlying fit was parameterised at BMD10, BMD20, or BMD50.
This means the parameterisation is not merely a computational convenience —
it changes the effective model weights and therefore the final regulatory
endpoint.

For practitioners who need BMD10 for regulatory purposes, a pragmatic
workflow would be:

1. Fit at BMD50 to obtain stable, well-identified model weights.
2. Derive BMD10 post-hoc from the BMD50 posterior via analytical inversion
   (as demonstrated in Section 4).
3. Compare with a direct BMD10 fit as a sensitivity check.

This approach leverages the numerical stability of the BMD50
parameterisation while still delivering the BMD10 endpoint required by
guidance.

## The non-informative prior is a secondary concern

The prior on BMD in the EFSA tool is approximately uniform on
$[0,\, 3 \cdot \max(\text{dose})]$ (PERT shape parameter $g = 0.0001$).
While this prior is technically data-adaptive and not invariant to
reparameterisation (see Section 2), in practice it is sufficiently flat that
it has negligible influence on the posterior when the data are informative.
The online tool already uses this non-informative default, and for typical
regulatory datasets with 4–5 dose groups the likelihood dominates.

The more consequential choices are the **parameterisation** (which BMR to
use as the sampled parameter) and the **inference method** (Laplace vs HMC),
both of which have a larger impact on the final BMD10 estimate than the
prior.

## Frequentist vs Bayesian: different methods, different answers

Conventional dose-response analysis derives a BMD from either the best-fitting
model or a frequentist model average. These are fundamentally different from
the Bayesian model-averaged BMD:

- **Bayesian BMA** averages the full posterior predictive distribution over
  models, weighted by marginal likelihoods (bridge sampling or Laplace).
- **Frequentist MA** averages point estimates across models, weighted by
  information criteria (AIC/BIC), with confidence intervals from inverse
  regression, the delta method, or bootstrap.

Even when the same set of dose-response shapes is used, these approaches
produce different BMD10 estimates because the weighting schemes, interval
construction, and background estimation differ. In the present example, the
four frequentist models (LL.4, W1.4, W2.4, EXD.3) yield BMD10 estimates
ranging from 0.18 to 0.64, but the AIC weights concentrate almost entirely
on W2.4 (weight ≈ 0.9999), so the frequentist model-averaged BMD10 is
essentially the W2.4 estimate (~0.18). By contrast, the Bayesian MA
estimates are 0.61 (Bridge Sampling) and 0.08 (Laplace) — neither close to
the frequentist result.

Several factors amplify the discrepancy:

1. **Model set.** The frequentist and Bayesian analyses use different model
   families. BMABMDR fits 16 models (8 shapes × {Normal, Lognormal});
   the `bmd` package uses `drc` model functions (LL.4, W1.4, W2.4, EXD.3,
   etc.). Even when shapes overlap conceptually, the parameterisations and
   estimation methods differ.
2. **Weighting scheme.** AIC weights can be highly concentrated — here a
   single model receives 99.99% of the weight. Bayesian posterior model
   probabilities based on marginal likelihoods tend to be more diffuse,
   spreading weight across multiple models. This makes the Bayesian MA
   more robust to a single model dominating but also more sensitive to
   poorly identified models inflating uncertainty.
3. **Model averaging method.** Even within the `bmd` package, the choice
   of averaging type matters. Options include "Kang" (weighted average of
   per-model BMD/BMDL), "Buckland" (weighted average with SE propagation),
   "bootstrap", and "curve" (averaging the dose-response curves directly).
   These can produce different MA-BMD values from the same set of models
   and weights.
4. **Background estimation.** Frequentist models estimate the background
   response as a free parameter per model (e.g. the lower asymptote in
   LL.4), which can differ from the Bayesian model-averaged background.
   Since BMD10 is defined as a 10% change relative to background, different
   background estimates yield different BMD10 values even when the curves
   are similar.

This methodological scatter is an unavoidable consequence of model
uncertainty. It poses a genuine challenge for practitioners who must derive
a single reference point for risk assessment. Current EFSA guidance
recommends the Bayesian model-averaged BMDL as the point of departure, but
practitioners should be aware that:

1. The numerical value depends on the inference method (Laplace vs HMC).
2. The numerical value depends on the BMR parameterisation (BMD10 vs BMD50).
3. The numerical value depends on the model set entering the averaging.
4. A frequentist analysis of the same data will generally give a different
   answer — and the frequentist result itself depends on the choice of
   models, weighting scheme, and averaging method.

Transparent reporting of these dependencies — rather than presenting a single
BMD as if it were uniquely determined — is essential for scientifically
defensible risk assessment.
