---
title: "Bug analysis: inconsistent prior.weights normalisation in full.laplace_MA()"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`full.laplace_MA()` accepts a `prior.weights` vector that lets users specify which
of the 16 candidate models should be included (1 = include, 0 = exclude). The vector
can also carry fractional values to express **informative prior model probabilities**
, for example, down-weighting certain model families before seeing the data.

The bug: **the vector is normalised in two different, mutually incompatible ways
at two different points in the same function**, so the final posterior model weights
are not proportional to what the user supplied.

---

## Where the two normalisations happen

### Normalisation 1 - inside `fun.w()` loop (line ~1542)

After all model-specific Laplace approximations have been computed, the raw
unnormalised Laplace approximation weights are combined with the prior weights:

```r
# fun_Laplace.R, near the end of the Laplace weight block
prior.weights = prior.weights/sum(prior.weights)          # <-- Normalisation 1
lpw = (prior.weights * w) / sum(prior.weights * w)
```

At this point `prior.weights` is divided by the **total sum of all elements**,
including the zeros. Because zeros contribute nothing to the sum, the non-zero
elements are rescaled to sum to 1.

For a uniform inclusion vector `c(1,1,1,1,0,0,...,0)` (4 models included):

```{r norm1}
prior.weights <- c(1, 1, 1, 1, rep(0, 12))

# Normalisation 1: divide by sum of ALL elements
prior.weights_norm1 <- prior.weights / sum(prior.weights)
prior.weights_norm1
# [1] 0.25 0.25 0.25 0.25  0    0  ...  0
# Included models each get prior weight 0.25  ✓
```

This is correct for the uniform case. When a user passes fractional weights,
however, the weights after normalisation are proportional to the original values,
which is the expected behaviour:

```{r norm1_fractional}
prior.weights_frac <- c(2, 1, 1, 1, rep(0, 12))

prior.weights_frac / sum(prior.weights_frac)
# E4_N gets twice the prior probability of each other included model  ✓
```

### Normalisation 2 - in `sampling_MA()` (line ~1322)

The `sampling_MA()` function performs an **identical** initial normalisation step,
but places it at the start of the weight-computation block rather than after model
fitting:

```r
# fun_Sampling.R, bridge-sampling weight block
p.weights = prior.weights / sum(prior.weights)            # <-- Normalisation 2
```

And in the convergence-fallback path:

```r
prior.weights.new = prior.weights
prior.weights.new[div.models] = 0
p.weights.new = prior.weights.new / sum(prior.weights.new) # <-- Normalisation 2b
```

Here the same formula is applied, but **before** any models are removed due to
fitting failure. Contrast this with `full.laplace_MA()`, where `prior.weights[k]`
is set to `0` *in-place* for models that fail to converge **before** the
normalisation step runs. The normalisation in `full.laplace_MA()` therefore
automatically absorbs failed models; the normalisation in `sampling_MA()` does not.

---

## Demonstrating the inconsistency

The core issue becomes visible with a simple numerical example. Suppose the user
specifies equal weights for all 16 models, but model 5 (`G4_N`) fails during
optimisation inside `full.laplace_MA()`.

```{r demo}
# Simulate what full.laplace_MA() does internally

# User supplies equal weights
prior.weights <- rep(1, 16)

# Model 5 fails; its weight is zeroed out inside the loop
prior.weights[5] <- 0

# Normalisation 1 (as coded in full.laplace_MA)
pw_laplace <- prior.weights / sum(prior.weights)
pw_laplace
# 15 models each receive 1/15 ≈ 0.0667
sum(pw_laplace)    # 1.0

# Now simulate sampling_MA: normalisation happens BEFORE model 5 is known to fail
prior.weights_orig <- rep(1, 16)
pw_sampling <- prior.weights_orig / sum(prior.weights_orig)
pw_sampling
# All 16 models each receive 1/16 = 0.0625
sum(pw_sampling)   # 1.0

# When model 5 later fails, sampling_MA does NOT re-normalise;
# it simply passes pw_sampling[5] = 0.0625 as the prior probability for a
# model that doesn't exist. The posterior model probabilities returned by
# bridgesampling::post_prob() will use prior_prob values that do NOT sum to 1
# over the surviving models.
surviving        <- pw_sampling[-5]       # the 15 active models
surviving_normed <- surviving / sum(surviving)
cat("bridge sampling prior for surviving models sums to:", sum(surviving), "\n")
# 0.9375, not 1.0 - i.e. 6.25% prior weight has silently disappeared
```

The consequence is that bridge sampling receives `prior_prob` values summing to
`1 - 1/16 = 0.9375` instead of 1.0. While `bridgesampling::post_prob()` re-normalises
internally, this means the user's original equal-weight intention is respected, but
the *divergent-model fallback* path (`prior.weights.new`) re-normalises a second
time, creating a subtle double-normalisation:

```{r double_norm}
# Divergent-model fallback in sampling_MA
prior.weights_new <- rep(1, 16)
prior.weights_new[5] <- 0                    # model 5 diverged

# Normalisation applied to NEW weights (already derived from original pw)
p.weights.new <- prior.weights_new / sum(prior.weights_new)
p.weights.new
# Each surviving model: 1/15 = 0.0667  ✓  (correct here)

# BUT p.weights was already computed from the ORIGINAL weights:
p.weights_orig <- rep(1, 16) / 16    # = 0.0625 each, including model 5

# The two are inconsistent: p.weights uses 16 models, p.weights.new uses 15.
# The initial lpwb is computed from p.weights (16 models),
# then lpwb.conv is computed from p.weights.new (15 models),
# but the final returned weights mix results from both computations.
```

---

## The inconsistency between `full.laplace_MA()` and `sampling_MA()` side-by-side

```{r comparison}
# Helper: show effective prior weight per model after normalisation

pw_user <- c(2, 1, 1, 1, rep(0, 12))   # user wants E4_N twice as likely

# What full.laplace_MA() does:
# Step 1: models with prior.weights == 0 are NEVER fitted (weight stays 0)
# Step 2: weights normalised AFTER fitting
pw_laplace_final <- pw_user / sum(pw_user)
round(pw_laplace_final[1:4], 4)
# E4_N: 0.40, IE4_N: 0.20, H4_N: 0.20, LN4_N: 0.20  ✓

# What sampling_MA() does:
# Step 1: weights normalised BEFORE fitting
pw_sampling_init <- pw_user / sum(pw_user)
round(pw_sampling_init[1:4], 4)
# E4_N: 0.40, IE4_N: 0.20, H4_N: 0.20, LN4_N: 0.20
# (Happens to match here for the simple case where no model fails)

# But if model 2 (IE4_N) fails inside sampling_MA:
# full.laplace_MA() re-normalises over {E4_N, H4_N, LN4_N} → 0.50, 0.25, 0.25
pw_user_after_fail <- pw_user
pw_user_after_fail[2] <- 0
pw_laplace_refail  <- pw_user_after_fail / sum(pw_user_after_fail)
round(pw_laplace_refail[c(1, 3, 4)], 4)
# E4_N: 0.50, H4_N: 0.25, LN4_N: 0.25  ← re-normalised correctly

# sampling_MA() does NOT re-normalise prior weights after a fit failure;
# it zeros the marginal likelihood contribution and relies on post_prob() to
# handle the remaining models - but p.weights passed still sum to 0.8, not 1.
pw_sampling_after_fail <- pw_sampling_init
pw_sampling_after_fail[2] <- 0
cat("sampling_MA prior sum after failure:", sum(pw_sampling_after_fail[c(1,3,4)]), "\n")
# 0.80 — not 1.0, inconsistent with full.laplace_MA's 1.0
```

---

## Impact summary

| Scenario | `full.laplace_MA()` | `sampling_MA()` |
|---|---|---|
| All models converge, equal weights | Correct | Correct |
| Some models fail, equal weights | Correct (re-normalises) | Minor error (prior sums < 1 before `post_prob` re-normalises) |
| Some models fail, **informative** weights | Correct | **Incorrect**: surviving prior probs do not preserve user's ratios |
| User passes `c(1,1,0,...,0)` | Correct | Correct (zeros excluded cleanly) |

The practical impact is small in most default-usage scenarios because
`bridgesampling::post_prob()` internally normalises its `prior_prob` argument.
However, when the user specifies **unequal, informative prior weights** and some
models diverge, the final posterior model weights in `sampling_MA()` will not
correctly reflect the specified prior model probabilities.

---

## Recommended fix

The fix is to adopt one consistent normalisation strategy. The cleanest approach
is to **defer normalisation until after model fitting** in both functions, normalising
only over models that actually produced a valid result.

```r
# Proposed unified approach (applies to both full.laplace_MA and sampling_MA)

# After fitting loop: zero out failed/diverged models
failed_idx <- which(converged == 0 | !models_fitted)
prior.weights[failed_idx] <- 0

# Normalise once over surviving models
if (sum(prior.weights) == 0) {
  stop("All candidate models failed to fit.")
}
prior.weights <- prior.weights / sum(prior.weights)

# Then compute posterior model probabilities
lpw <- (prior.weights * w) / sum(prior.weights * w)
```

This makes the normalisation logic identical across both functions, ensures the user's
relative weighting intentions are preserved after failures, and removes the current
asymmetry between the two code paths.
